---
title: "Lab4"
author: "Skyler Moomey"
date: "'r Sys.Date()'"
output: 
html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tasks

## Task 1
```{r}
getwd()
```

## Task 2
```{r}
#spruce data frame
spruce.df <- read.csv("SPRUCE.csv")
names(spruce.df)[1] = "BHDiameter"

tail(spruce.df)
```

## Task 3
```{r}
library(s20x)

#lowess plot
trendscatter(Height~BHDiameter,f=0.5,data=spruce.df)
# Now make the linear model
spruce.lm=lm(Height~BHDiameter,data=spruce.df)
summary(spruce.lm)

height.res=residuals(spruce.lm) #residuals
height.fit=fitted(spruce.lm) #fitted values

#plot of residuals v. fitted
trendscatter(height.res~height.fit, f=0.5)
```
The residuals vs. fitted value make a very messy plot, and the trendscatter function tries to smooth them into a roughly quadratic shape, even though it barely fits the data.

```{r}
with(data=spruce.df, plot(x=BHDiameter, y=Height, pch=21, ylab="Height of Tree(m)", xlab="Diameter at Breast Height (cm)", bg="blue", 
                          ylim=c(0, 1.1*max(Height)), xlim=c(0, 1.1*max(BHDiameter)), cex=1.2))
title(main="Regression Plot with Residuals")
yhat=with(spruce.df, predict(spruce.lm, data.frame(BHDiameter)))
with(data=spruce.df, segments(BHDiameter, Height, BHDiameter, yhat))
abline(spruce.lm, col="dark green")

normcheck(spruce.lm, shapiro.wilk = TRUE)
title(main="Roughly Normal Distribution")
mean(height.res)
I(sd(height.res))^2
```
The p-value $=0.29$ for the linear model. This is much higher than the necessary $0.05$ to reject the null hypothesis.
Additionally, the mean of the residuals is approximately $-3.09*10^{-17}$, extremely close to 0. $\sigma^2=2.734$ and so the linear model is validly applied to the SPRUCE data set, though there are still better models.

## Task 4
```{r}
## Quadratic object using the linear model
quad.lm=lm(Height~BHDiameter + I(BHDiameter^2),data=spruce.df)
quad.res=residuals(quad.lm)

with(spruce.df, plot(Height~BHDiameter, main="Quadratic Regression", pch=21, cex=1.2, ylim=c(0, 1.1*max(Height)), xlim=c(0, 1.1*max(BHDiameter)),
                                                                                                                         bg="light blue"))
#function to add quadratic curve to scatterplot
myplot=function(x){
 quad.lm$coef[1] +quad.lm$coef[2]*x  + quad.lm$coef[3]*x^2
} 

curve(myplot, lwd=2, add=TRUE, col="purple")

plot(quad.lm, which=1)

normcheck(quad.lm, shapiro.wilk= TRUE, main="Normal Distribution Check")
mean(quad.res)
I(sd(quad.res))^2

```
The quadratic model's p-value is 0.684, which allows us to accept the null hypothesis. The mean of the residuals is once again close to $0$ $\mu=-3.47*10^{-17}$, so the model is again a valid estimator of the data.
